Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-06-01 ms.tex
small changes
filling out transitions in introduction section

\newcommand{\be}{\begin{eqnarray}}
\newcommand{\ee}{\end{eqnarray}}
\def\la{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'074$}}\mathchar"7218$}}}
\def\ga{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'076$}}\mathchar"7218$}}} %> or of order
\renewcommand{\vec}[1]{\mathbf{#1}}
%\renewcommand{\vec}[1]{\ensuremath{\boldsymbol{#1}}} %boldface vector style
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\proposed}[1]{\textcolor{blue}{#1}}

%%% FOR DRAFTING: REMOVE AFTER PAPER IS COMPLETE!
\usepackage{xcolor}

@@

\proposed{
The outer layers of planetary atmospheres are central to a planet's evolution, as they can shelter the lower atmosphere from high energy radiation and regulate the escape of gas into space.} % Add something about lyman alpha here
%There's a much thicker layer of atomic hydrogen
%It's a very optically thick region
% Care about lya going into the planet's atmosphere and dissociating molecules
% Observational: observe the transits in lyman alpha
%Chen liang's paper - focus on the section where he shows the results of the lya transfer - can cause heating deeper in the atmosphere, collisional de-excitation. Summarize some of the things that lya does to the atmosphere. Ionize low ionization-potential atoms (<10.2 eV)
%Large Lya line intensity, lots of H atoms, then lya is constantly exciting and de-exciting. Then, n=2 is populated so Balmer lines can also be absorbed. This explains why strong balmer absorption is observed in transits to begin with. Otherwise, you would need very high temperatures to populate the n=2 state.
% Focus on later sections, especially balmer absorption near the end of the paper.
\todo{There exist several fundamental questions about the evolution and stability of exoplanet atmospheres that have become irradiated by their host stars; understanding which physical and chemical processes are important in these regions can assist in the creation of robust models to answer such questions. (TODO: This sentence reads mostly like fluff but I'm not sure how to outline the problem otherwise.)}

@@

% Put emphasis on the fact that Lya is generated within the atmosphere --- ionization and rate equilibrium balanced by recombination. Recombination gives you lyman alpha again. Collisions decrease the density in the 2s states. The efficiency of producing lya is quite high, and that's immediately at high optical depth.

\proposed{
Hubble Space Telescope (HST) observations with STIS have found large Lyman $\alpha$ transit depths around the gas giants HD 209458b \citep{2003Natur.422..143V}, HD 189733b \citep{2012A&A...543L...4L} and 55 Cnc b \citep{2012A&A...547A..18E}, Neptune-size planets GJ 436 b
\citep{2015Natur.522..459E}
and GJ 3470 b \citep{2018A&A...620A.147B}, and possibly the super-Earth HD 219134b \citep{2019EPSC...13.1928L}. In the case of GJ 436 b, ultraviolet transit depths of 56\% are reported, claiming a total Ly-$\alpha$ eclipse would have been possible for a central transit. The most notable absorption is stated to occur in the blue wing of the Ly-$\alpha$ line. Follow-up observations of GJ 3470 b by \citet{2021arXiv210309864B} confirm the presence of an exospheric hydrogen cloud around the planet with a radius of 1.3 times its Roche lobe in the far-ultraviolet, implying that H atoms are escaping the atmosphere. 
Non-detections or marginal detections have been reported for the super-Earths 55 Cnc e \citep{2012A&A...547A..18E}, super-Earth HD 97658 b \citep{2017A&A...597A..26B} , super-Earth GJ 1132 b \citep{2019AJ....158...50W}, super-Earth $\pi$ Men c \citep{2020ApJ...888L..21G}, and marginal detections were Earth-size TRAPPIST-1 \citep{2017A&A...599L...3B} and sub-Earth Kepler-444 \citep{2017A&A...602A.106B}.  \cite{2021ApJ...907L..36M} reports a detection of CII ions in a single transit of the exoplanet $\pi$ Men c, with transit depths and Doppler velocities consistent with an escaping atmosphere.
}

\proposed{
These observations have revealed a population of atoms extending out to distances of order the planetary radius for several planets around bright, nearby stars. In the case of HD 209458b, observations detected atomic hydrogen (H) absorption in the stellar Ly-$\alpha$ line during three transits. Observational data was taken from the Space Telescope Imaging Spectrograph (STIS) onboard the Hubble Space Telescope (HST) using the G140M grating to minimize contamination from Earth's Ly-$\alpha$ geocoronal emission. The planet's radius and mass suggest the absorption takes place beyond the Roche limit (i.e., the absorption is occurring in escaped H atoms). The magnitude of absorption (15\%) is larger than it should be for a planet that occults only 1.5\% of the star's visible surface. A maximally-filled Roche lobe would account for only 10\% absorption. The outlying 5\% suggests that hydrogen is escaping the planet's atmosphere. Additionally, the transits show blue-shifted absorption, possibly due to acceleration of the planet’s atmosphere by radiation pressure from the star. Models designed by \citet{2003Natur.422..143V} suggest that the escaping hydrogen atoms expand in an asymmetric comet-like tail and progressively disappear when moving away from the planet, which is consistent with most observations mentioned here. The absorption is thought to last well after the end of the transit due to the large physical size of the evaporating tail. In this way, planetary mass and radius measurements allow detailed analysis of Ly-$\alpha$ spectral features, as it is possible to discern what fraction of the absorbing atoms are gravitationally bound to the planet.
}

\todo{Transition to semi-analytic approach, justification for building models}

@@

\citet{1973MNRAS.162...43H, 1974MNRAS.166..373H} demonstrated that resonance-line radiation transfer in low-density, optically-thick media can be described by the Poisson equation. However, the accuracy of their solution is limited by the treatment of the boundary conditions. In their eigenfunction expansion, the spatial and frequency components of the differential equation are separated, yet the spatial eigenvalues are defined as being explicitly frequency-dependent. The resulting solution for large optical depths would suggest the mean intensity at the boundary of the slab is zero, which is unphysical. A more detailed discussion of the boundary condition is included in Section \ref{sec:no_destruction}. \citet{1990ApJ...350..216N} follows this approach in an attempt to extend the result to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line, but again does not use the correct boundary condition. \citet{1990ApJ...350..216N} improved on this approach, producing a divergent, inhomogeneous solution to the problem in 1D slab geometry that includes the effect of photon absorption within the medium. \citet{2006ApJ...649...14D} generalize the problem to spherical geometry, which is the same physical set up used here. Following this, \citet{2020arXiv200509692L} presents closed-form solutions for the radiation field in spherical geometry, utilizing the ``core skipping'' method discussed in \citet{2015MNRAS.449.4336S} to handle photon frequency redistribution within the Doppler core. This technique involves drawing the atom's velocity from a truncated Gaussian distribution, forcing photons back into the wing when they scatter into the core. This is motivated by the assumption that all core scattering results in zero spatial diffusion until the photon is redistributed back out into the wing where the mean free path becomes much larger and it can diffuse in space again. \todo{However, it is not clear whether this prescription is fully independent of the choice of velocity distribution.}

>>>>

\newcommand{\be}{\begin{eqnarray}}
\newcommand{\ee}{\end{eqnarray}}
\def\la{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'074$}}\mathchar"7218$}}}
\def\ga{\mathbin{\lower 3pt\hbox
      {$\rlap{\raise 5pt\hbox{$\char'076$}}\mathchar"7218$}}} %> or of order
\renewcommand{\vec}[1]{\mathbf{#1}}
%\renewcommand{\vec}[1]{\ensuremath{\boldsymbol{#1}}} %boldface vector style
\newcommand{\grad}{\mathbf{\nabla}}
\newcommand\lya{Ly$\alpha$\ }

%%% FOR DRAFTING: REMOVE AFTER PAPER IS COMPLETE!
\usepackage{xcolor}
\newcommand{\todo}[1]{\textcolor{red}{#1}}
\newcommand{\proposed}[1]{\textcolor{blue}{#1}}

@@

The outer layers of planetary atmospheres are central to a planet's evolution, since they can shelter the lower atmosphere from high energy radiation and regulate the escape of gas into space. In atmospheres with thick layers of atomic hydrogen, stellar Lyman Alpha (Ly$\alpha$) radiation is especially important as the optical depth at line center is large. Ionization and subsequent recombination can generate \lya within the planet's atmosphere, dissociating molecules in lower layers and causing heating by collisional de-excitation. It can also excite H atoms to the 2p state, creating a population of Balmer-line absorbers that can be observed via transmission spectroscopy.

%There's a much thicker layer of atomic hydrogen
%It's a very optically thick region
% Care about lya going into the planet's atmosphere and dissociating molecules
% Observational: observe the transits in lyman alpha
%Chen liang's paper - focus on the section where he shows the results of the lya transfer - can cause heating deeper in the atmosphere, collisional de-excitation. Summarize some of the things that lya does to the atmosphere. Ionize low ionization-potential atoms (<10.2 eV)
%Large Lya line intensity, lots of H atoms, then lya is constantly exciting and de-exciting. Then, n=2 is populated so Balmer lines can also be absorbed. This explains why strong balmer absorption is observed in transits to begin with. Otherwise, you would need very high temperatures to populate the n=2 state.
% Focus on later sections, especially balmer absorption near the end of the paper.

@@

% Put emphasis on the fact that Lya is generated within the atmosphere --- ionization balanced by recombination, rate equilibrium. Recombination gives you lyman alpha again. Collisions decrease the density in the 2s states. The efficiency of producing lya is quite high, and that's immediately at high optical depth.

Hubble Space Telescope (HST) observations with STIS have found large Lyman $\alpha$ transit depths around the gas giants HD 209458b \citep{2003Natur.422..143V}, HD 189733b \citep{2012A&A...543L...4L} and 55 Cnc b \citep{2012A&A...547A..18E}, Neptune-size planets GJ 436 b
\citep{2015Natur.522..459E}
and GJ 3470 b \citep{2018A&A...620A.147B}, and possibly the super-Earth HD 219134b \citep{2019EPSC...13.1928L}. In the case of GJ 436 b, ultraviolet transit depths of 56\% are reported, claiming a total Ly-$\alpha$ eclipse would have been possible for a central transit. Follow-up observations of GJ 3470 b by \citet{2021arXiv210309864B} confirm the presence of an exospheric hydrogen cloud around the planet with a radius of 1.3 times its Roche lobe in the far-ultraviolet, implying that H atoms are escaping the atmosphere. 
Non-detections or marginal detections have been reported for the super-Earths 55 Cnc e \citep{2012A&A...547A..18E}, super-Earth HD 97658 b \citep{2017A&A...597A..26B} , super-Earth GJ 1132 b \citep{2019AJ....158...50W}, super-Earth $\pi$ Men c \citep{2020ApJ...888L..21G}, and marginal detections were Earth-size TRAPPIST-1 \citep{2017A&A...599L...3B} and sub-Earth Kepler-444 \citep{2017A&A...602A.106B}.  \cite{2021ApJ...907L..36M} reports a detection of CII ions in a single transit of the exoplanet $\pi$ Men c, with transit depths and Doppler velocities consistent with an escaping atmosphere.

These observations have revealed a population of atoms extending out to distances of order the planetary radius for several planets around bright, nearby stars. The importance of the \lya line in the atmospheric dynamics of these systems necessitates a careful treatment of resonance-line radiation transfer in order to construct accurate models of atmospheric heating and escape. However, the extreme optical depths at \lya line center introduce a computational barrier to simulating radiative transfer via Monte Carlo methods.

@@

\citet{1973MNRAS.162...43H, 1974MNRAS.166..373H} demonstrated that resonance-line radiation transfer in low-density, optically-thick media can be described by the Poisson equation. However, the accuracy of their solution is limited by the treatment of the boundary conditions. In their eigenfunction expansion, the spatial and frequency components of the differential equation are separated, yet the spatial eigenvalues are defined as being explicitly frequency-dependent. The resulting solution for large optical depths would suggest the mean intensity at the boundary of the slab is zero, which is unphysical. A more detailed discussion of the boundary condition is included in Section \ref{sec:no_destruction}. \citet{1990ApJ...350..216N} follows this approach in an attempt to extend the result to media of intermediate optical depth, including the effects of scattering in the Doppler core of the line, but again does not use the correct boundary condition. \citet{?} improved on this approach, producing a divergent, inhomogeneous solution to the problem in 1D slab geometry that includes the effect of photon absorption within the medium. \citet{2006ApJ...649...14D} generalize the problem to spherical geometry, which is the same physical set up used here. Following this, \citet{2020arXiv200509692L} presents closed-form solutions for the radiation field in spherical geometry, utilizing the ``core skipping'' method discussed in \citet{2015MNRAS.449.4336S} to handle photon frequency redistribution within the Doppler core. This technique involves drawing the atom's velocity from a truncated Gaussian distribution, forcing photons back into the wing when they scatter into the core. This is motivated by the assumption that all core scattering results in zero spatial diffusion until the photon is redistributed back out into the wing where the mean free path becomes much larger and it can diffuse in space again. \todo{However, it is not clear whether this prescription is fully independent of the choice of velocity distribution.}
